---
title: "Transmission type effects on fuel economy"
author: "Alex Gaggin"
date: "Thursday, October 25, 2015"
output:
  pdf_document

---

### Summary

While intuitively it looks like manual transmission provides much better fuel
economy, statistical analysis of specifications for 32 cars of 1973-74 year
models shows that most of this effect caused by manufacturers' preference
to install automatic transmissions on heavier cars, which are naturally driven
less miles per gallon. Transmission type does influence fuel economy, but it's
third source of fuel economy influence after weight and quarter mile time. As
it's trumped by first two influencers, its size of effect is hard to measure
confidently - the data shows that **switching to manual can save from 0.05 to
5.8 miles per gallon**.


```{r, include=FALSE}
# Load data and libraries
library(dplyr)
library(datasets)
data(mtcars)
```

### Exploration of the Motor Trends dataset

Dataset has 32 observations on 11 variables and their descriptions are available
in dataset's help page.
By checking correlations, we see that weight variable affects the fuel economy
the most - and by plotting mpg and weight against transmission types (see
Appendix 1) we see that weight difference for transmissions is stronger than mpg
difference.

```{r}
cor(mtcars)[, "mpg"] %>% "["(-1) %>% abs %>% sort(decr=TRUE) %>% signif(2)
```

### Significance of MPG difference for transmission types

```{r}
t.test(mtcars[mtcars$am == 0, ]$mpg, mtcars[mtcars$am == 1, ]$mpg)$p.value
```

Small p-value shows that we have to accept alternative hypothesis and presume
there's a meaningful difference in mean mpgs between automatic and manual
transmissions. Still, strong relation doesn't mean causation. Let's
model if this relation still stands after other factors are taken into account.

### Linear models

First, let's model dependance just for weight, and then find which of other
variables is most correlated with the residuals. We add then this variable
to the model as a second regressor.

```{r}
fit1 <- lm(mpg ~ wt, mtcars)
mtcars[ ,-c(1, 6)] %>% cbind(resid(fit1)) %>% cor %>%
        "["(, 10) %>% "["(1:9) %>% abs %>% which.max %>% names
fit2 <- lm(mpg ~ wt + qsec, mtcars)
summary(fit2)$coefficients
```

We see that time to quarter mile is also significant. It makes sense -
presumably
there should be two types of wasteful cars - large (heavy) ones,
and sporty (with rapid acceleration to quarter mile) ones.  
Reiterating the same
logic, we build next two nested models.

```{r, include=FALSE,eval=FALSE}
mtcars[ ,-c(1, 6, 7)] %>% cbind(resid(fit2)) %>% cor %>%
        "["(, 9) %>% "["(1:8) %>% abs %>% which.max %>% names
mtcars[ ,-c(1, 6, 9, 7)] %>% cbind(resid(fit3)) %>% cor %>%
        "["(, 8) %>% "["(1:7) %>% abs %>% which.max %>% names
```
```{r}
fit3 <- lm(mpg ~ wt + qsec + factor(am), mtcars)
fit4 <- lm(mpg ~ wt + qsec + factor(am) + carb, mtcars)
anova(fit1, fit2, fit3, fit4)
```

We see that three-variable model still significantly improved results, while
including carb variable doesn't improve it further. Residuals for three-variable
model don't have any apparent patterns in them (see Appendix 2), although they aren't ideally
normal - a bit skewed. Still, we accept this model as the best linear model we
could build on this dataset.

```{r}
fit3 %>% confint %>% signif(2)
```

We see that manual transmissions allow to get better mileage, although exact
scale of that effect is hard to pin down, as it's trumped by bigger effects
of cars' weights and acceleration capabilities expressed as the quarter mile time.
It's barely fits 95% significance rule and can be anywhere between 0.05 and 5.8
miles per gallon.

\newpage

## Appendices

### Appendix 1. Charts for mpg and weight dependance on transmission types

```{r}
par(mfrow = c(1, 2))
amFactor <- factor(mtcars$am, levels=0:1, labels=c("Automatic", "Manual"))
plot(amFactor, mtcars$mpg, xlab="Transmission type", ylab="Miles per gallon")
plot(amFactor, mtcars$wt, xlab="Transmission type", ylab="Weight, 1000lb")
```

\newpage

### Appendix 2. Residuals analysis for the selected model

```{r}
par(mfrow = c(1, 2)); plot(fit3, which = 1:2)
```

